TrenchR: an R package for transparent environmental and ecological biophysics

### 2022-06-14 Manuscript_full.Rmd
minor edits

```{r, echo=FALSE, fig.height=4, out.width = "75%"}
knitr::include_graphics('figures/TrenchRBiophys.png')
```

@@

The TrenchR package is intended to complement the NicheMapR package,
which includes sophisticated microclimate [@kearney2017nichemapr] and
biophysical [@kearney2020nichemapr] models. Although the Fortran source
code was recently released, the complexity of NicheMapR functions can
make it difficult to understand and modify the source code. Many of the
approaches are similar between the packages, but TrenchR provides simple
functions to aid the understanding and transparency of biophysical
approaches. Some variable definitions and units are lacking in NicheMapR. The TrenchR functions are modular and easily adapt to a
variety of organisms and research questions. Although many of the
TrenchR functions are general, some components of TrenchR are best
suited for ectothermic animals. We omit plant specific biophysical
models since they are the focus of several R packages [tealeaves:
[@muir2019tea]; plantecophys: [@duursma2015plantecophys]]. Some energy
budget calculation components of biophysical models are also available
in the ThermImage package [@tattersall2017]. Microclimate models, with
an emphasis on describing spatial variation, are also available in the
Microclima package [@maclean2019microclima].

@@

We adapted functions from biophysical ecology textbooks
[@campbell2000introduction; @gates1980biophysical] and research
articles. Source references are included in function headers. We have
incorporated default parameters (e.g., organism emissivity or heat
conduction rates) when general values are available. Many functions
additionally include comments that describe potential parameterizations.
The biophysical ecology texts [@campbell2000introduction;
@gates1980biophysical] include many tables with parameter values and we
provide several data tables in TrenchR.

@@

TrenchR functions are organized into the following categories:

@@

@@

@@

For additional background on microclimate and biophysical modelling, we
have updated a series of tutorials entitled *Physical Processes in
Ecosystems*. We intend the tutorials to provide a more contained and
accessible introduction to microclimate and biophysical modelling than
that included in classic textbooks [@campbell2000introduction;
@gates1980biophysical]. The tutorials provide less detailed coverage of
many of the topics contained in the books. The series of 14 tutorials
starts with an overview of the calculus and physics principles
underlying the modelling. Thermodynamics and energy budget modelling are
then introduced with detailed examples. Tutorials address the climate
space concept, operative temperatures, biophysical models for leaves and
sheep, and heat flow in soils. We have revised the tutorials (which
originated from an NSF training grant in 1979 and lacked broad
distribution) to include R code and utilization of TrenchR functions. We
expanded the original series to include tutorials contributed by M.
Kearney introducing the Microclim environmental data
[@kearney2014microclim] and integrating the Microclim data with
biophysical modelling to examine broad scale climatic limits. The
tutorials are provided in html form and available for download as a pdf
at the bookdown server
(<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>). The
R markdown files are available in Github
(<https://github.com/trenchproject/TrenchRmodules>).

## Examples

We illustrate the use of the TrenchR package by estimating an energy
budget for a *Sceloporus* lizard on June 1, 2021 in Santa Fe, New
Mexico, USA (35.69, -105.944, elevation: 2121 meters). The example is
also incorporated in the *Using energy balances to estimate body
temperatures* vignette. We start by generating environmental inputs.
Using these inputs, we estimate the energy budget with component
functions. Finally, we use an integrated biophysical model to estimate
operative environmental temperatures. See Appendix S1 for another
example examining a time series of microclimate data and applying and
testing the biophysical models.
**Flow cart of functions?

>>>>

```{r, echo=FALSE, fig.height=4, out.width = "75%"}
knitr::include_graphics('./figures/TrenchRBiophys.png')
```

@@

The TrenchR package is intended to complement the NicheMapR package,
which includes sophisticated microclimate [@kearney2017nichemapr] and
biophysical [@kearney2020nichemapr] models. Although the Fortran source
code was recently released, the complexity of NicheMapR functions can
make it difficult to understand and modify the source code. The function documentation including variable definitions and units can also be challenging to follow. Many of the
approaches are similar between the packages, but TrenchR provides simple
functions to aid the understanding and transparency of biophysical
approaches. The TrenchR functions are modular and easily adapt to a
variety of organisms and research questions. Although many of the
TrenchR functions are general, some components of TrenchR are best
suited for ectothermic animals. We omit plant specific biophysical
models since they are the focus of several R packages [tealeaves:
[@muir2019tea]; plantecophys: [@duursma2015plantecophys]]. Some energy
budget calculation components of biophysical models are also available
in the ThermImage package [@tattersall2017]. Microclimate models, with
an emphasis on describing spatial variation, are also available in the
Microclima package [@maclean2019microclima].

@@

We adapted functions from biophysical ecology textbooks
[@campbell2000introduction; @gates1980biophysical] and research
articles. Source references are included in function headers and reference sections. We have
incorporated default parameters (e.g., organism emissivity or heat
conduction rates) when general values are available. Many functions
additionally include comments that describe potential parameterizations.
The biophysical ecology texts [@campbell2000introduction;
@gates1980biophysical] include many tables with parameter values and we
provide several data tables in TrenchR.

@@

TrenchR functions are organized into the following categories (Figure 2):

@@

**Allometric functions** allow estimating the dimensions of organisms needed for energy balances and other analyses. The functions can convert between organismal length, mass, surface area, and volume. The functions can also estimate the silhoutte area, which describes the organismal area exposed to solar radiation.

@@

```{r, echo=FALSE, fig.height=4, out.width = "75%"}
knitr::include_graphics('./figures/TrenchRFlow.png')
```

**Figure 2** An examplar aim of translating from environmental and phenotypic data to organismal conditions in a given environment (grey boxes) can be achieved in TrenchR via several categories of functions (white boxes). We list example functions for each category used in the example application below.

@@

For additional background on microclimate and biophysical modelling, we
have updated a series of tutorials entitled *Physical Processes in
Ecosystems*. We intend the tutorials to provide a more contained and
accessible introduction to microclimate and biophysical modelling than
that included in classic textbooks [@campbell2000introduction;
@gates1980biophysical]. The tutorials provide less detailed coverage of
many of the topics contained in the books. The series of 14 tutorials
starts with an overview of the calculus and physics principles
underlying the modelling. Thermodynamics and energy budget modelling are
then introduced with detailed examples. Tutorials address the climate
space concept, operative temperatures, biophysical models for leaves and
sheep, and heat flow in soils. We have revised the tutorials (which
originated from an NSF training grant in 1979 and lacked broad
distribution) to include R code and utilization of TrenchR functions. We
expanded the original series to include tutorials contributed by M.
Kearney introducing the Microclim environmental data
[@kearney2014microclim] and integrating the Microclim data with
biophysical modelling to examine broad scale climatic limits. The
tutorials are provided in html form and available for download as a pdf
at the bookdown server
(<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>).
R markdown files for the tutorials are available in Github
(<https://github.com/trenchproject/TrenchRmodules>).

## Example application

We illustrate the use of the TrenchR package by estimating an energy
budget for a *Sceloporus* lizard on June 1, 2021 in Santa Fe, New
Mexico, USA (35.69°, -105.944°, elevation: 2121 m). The example is
also incorporated in the *Using energy balances to estimate body
temperatures* vignette. We start by generating environmental inputs (Figure 2).
Using these inputs, we estimate the energy budget with component
functions. Finally, we use an integrated biophysical model to estimate
operative environmental temperatures. See Appendix S1 for another
example examining a time series of microclimate data and applying and
testing the biophysical models.
